---
title: "Data Preparation for Lao AMR Dashboard"
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_file = file.path('./data/', paste0("Data_Preparation_AMR_", format(Sys.time(), '%Y-%m-%d_%Hh%M'), ".html"))) })
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

rm(list = ls())
library(lubridate)
library(readxl)
library(tidyverse)

time_start <- format(Sys.time(), '%Y-%m-%d_%Hh%M')
```

# About, How to Use

The `Data_Preparation.Rmd` script is a R Markdown document that aims to generate a data file ready for use in the Lao AMR Dashboard.
*R Markdown is a file format for making dynamic documents with R. An R Markdown document is written in markdown (an easy-to-write plain text format) and contains chunks of embedded R code. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. To learn more about R Markdown, [visit RStudio website](https://rmarkdown.rstudio.com/index.html)*

**In order to run ("Knit") this script, one will need first to install software:**

- Install R Statistical software that can be downloaded from [The Comprehensive R Archive Network (CRAN)](https://cran.r-project.org).
- Install RStudio Desktop (Open Source Licence), an integrated development environnement for R that can be downloaded from [RStudio website](https://www.rstudio.com/products/rstudio/download/).
- Install the following packages: `lubridate`, `markdown`, `readxl`, `tidyverse` by click on the Packages tab in the bottom-right section and then click on Install.

The next thing to do is to ensure that the data is available as two Excel files located in a folder named `data` and which is a child folder from this script.

Once all required elements are available and [the user "Knit" this R Markdown file](https://rmarkdown.rstudio.com/lesson-2.html) successfully, two files are generated:

- `Data_Preparation.html` that describes several steps and highlight important elements of the preparation of the data.
- `AMR_data_LOMWRU_Latest.RData` with the data ready to use in the dashboard.

One potential source of errors lie in setting up the working directory in R.

## Import data

To successfully import the data, files should be located in a folder named `data` and which is a child folder from this script.

**Please carefully review possible message generated during the import process.**

```{r, echo = TRUE}
# "./" refers to the device working directory — the R command getwd() provides the current working directory
# when executing the command read_excel("./data/LIMS_AMR_2018-10-03.xlsx", guess_max = 21474836)
# R will try to import the Excel file "LIMS_AMR_2018-10-03.xlsx" that is expected to be located in a
# data subfolder in the current working directory

if (interactive()) setwd("./Lao-AMR-Dashboard")
getwd()

amr <- read_excel("./data/LIMS_AMR_01082019.xlsx", guess_max = 21474836)
antibio_codes <- read_excel("./data/LIMS_AMR_codes.xlsx", sheet = 1)
spec_codes <- read_excel("./data/LIMS_AMR_codes.xlsx", sheet = 2)
org_include <- read_excel("./data/LIMS_AMR_codes.xlsx", sheet = 3)
```

```{r eval = FALSE, echo = FALSE}
# NOT RUN
# List of comments generated for manual filter by researchers
write.csv(sort(unique(c(amr$culturecomment, amr$culturecomment2))), file = "all_comments_amr.csv")
```

The following columns are missing from the dataset:

```{r}
setdiff(c("specnum", "patient_idnum", "name", "surname", "dob", "gender", 
"villagecode", "Village", "districtcode", "District", "provincecode", 
"Province", "patientID_01", "spectype", "specdate", "age_years", 
"age_months", "age_days", "location", "ward", "specdetail", "speccomment1", 
"org_code", "ORG_NAME", "ORG_GROUPNAME", "orgnum", "culture_date", 
"orgquant", "idtype", "idprofile", "idpercent", "esbl", "caz", 
"cazcv", "cpd", "cpdcv", "ctx", "ctxcv", "serotype", "culturecomment", 
"culturecomment2", "culturenote", "Otherinfo_PACsID", "Otherinfo_straincode", 
"TestType", "Antibiotic_ETEST_name", "Measurement", "Resistance", 
"Note"), names(amr))
```

The following columns were not required and will be disregarded during analysis:

```{r}
setdiff(names(amr), c("specnum", "patient_idnum", "name", "surname", "dob", "gender", 
"villagecode", "Village", "districtcode", "District", "provincecode", 
"Province", "patientID_01", "spectype", "specdate", "age_years", 
"age_months", "age_days", "location", "ward", "specdetail", "speccomment1", 
"org_code", "ORG_NAME", "ORG_GROUPNAME", "orgnum", "culture_date", 
"orgquant", "idtype", "idprofile", "idpercent", "esbl", "caz", 
"cazcv", "cpd", "cpdcv", "ctx", "ctxcv", "serotype", "culturecomment", 
"culturecomment2", "culturenote", "Otherinfo_PACsID", "Otherinfo_straincode", 
"TestType", "Antibiotic_ETEST_name", "Measurement", "Resistance", 
"Note"))
```


## Selection and rename of columns

```{r}
amr <- amr %>%
  select(spec_id = specnum,
         patient_id = patient_idnum,
         patient_name = name, # surname, dob,
         gender,
         # villagecode, Village, districtcode, District, provincecode,
         province = Province,
         # patientID_01,
         spec_type = spectype,
         spec_date = specdate,
         age_years, age_months, age_days,
         location,
         # ward, specdetail, speccomment1, org_code,
         org_name = ORG_NAME,
         # ORG_GROUPNAME, orgnumm, culture_date, orgquant, idtype, idprofile, idpercent,
         esbl,
         # caz,	cazcv, cpd, cpdcv, ctx, ctxcv, serotype, 
         culturecomment, culturecomment2, 
         # culturenote, Otherinfo_PACsID, Otherinfo_straincode, TestType,
         antibiotic_code = Antibiotic_ETEST_name,
         # Measurement,
         resistance = Resistance
         # Note
  ) %>% 
  mutate(
    spec_year = year(spec_date),
    antibiotic_code = str_to_upper(antibiotic_code))

antibio_codes <- antibio_codes %>%
  select(antibiotic_code, 
         method = Method,
         antibiotic_name)

spec_codes <- spec_codes %>%
  select(spec_type, 
         spec_method)
```


```{r}
# Include antibiotic names from the antibiotic codes.
amr <- left_join(amr, antibio_codes, by = "antibiotic_code")
```


```{r}
# Include specimen names from the specimen codes.
amr <- left_join(amr, spec_codes, by = "spec_type")
```

The `amr` dataset contains `r nrow(amr)` rows. There are `r length(unique(amr$patient_id))` patients  and `r length(unique(amr$spec_id))` specimens.

The `amr` dataset contains the following columns:

- `spec_id`: specimen id
- `patient_id`: patient id
- `gender`: sex of the patient
- `province`: province of residence of the patient
- `spec_type`: code of the type of specimen (blood culture, ear swab...)
- `spec_date`: date of collection of the specimen
- `age_years`, `age_months`, `age_days`: age of the patient
- `location`: location (hospital) of the specimen collection
- `org_name`: name of the organism
- `esbl`: ESBL for some organisms (Escherichia coli ...)
- `antibiotic_code`: code of the antibiotic
- `resistance`: resistance status
- `method`: Disc or Etest
- `antibiotic_name`: name of the antibiotic
- `spec_method`: method of collection of the specimen


## Data Quality Check

### Recode values

The following columns have missing values:

```{r}
# check columns with missing values
missing <- function(x) any(is.na(x))
missing_columns <- apply(amr, 2, missing)
names(missing_columns[missing_columns == TRUE])
```

```{r echo = TRUE}
amr <- amr %>% 
  replace_na(list(gender = "~ Unknown", 
                  province = "Xaysomboun", 
                  location = "~ Unknown", 
                  org_name = "~ Unknown", 
                  esbl = "Negative",  # Following Tamalaee instructions. Negative test results should be made explicit in future updates.
                  spec_method = "~ Unknown",
                  culturecomment = "-",
                  culturecomment2 = "-"))
```

Recode location (OPD Mahosot, Police and Military Hospitals).
```{r echo = TRUE}
amr$location[amr$location == "Mahosot Hospital"] <- "Mahosot"
amr$location[amr$location == "Mahosot hospital"] <- "Mahosot"
amr$location[amr$location == "OPD"] <- "Mahosot"
amr$location[amr$location == "Sethathirat Hospital (ST)"] <- "Setthathirat Hospital (ST)"
amr$location[amr$location == "Police Hospital (PH)"] <- "Police Hospital Vientiane"
amr$location[amr$location == "Military Hospital (MIL)"] <- "Military Hospital Vientiane (MIL)"
amr$location[amr$location == "Xieng Kwang Hospital (XK)"] <- "Xieng Khuang Hospital (XK)"
amr$location[amr$location == "Luxembourg hospital, Vientiane Province (VT)"] <- "Vientiane Provincial Hospital (VTP)"
```


Recode province names.
```{r echo = TRUE}
amr <- amr %>% 
  mutate(
  province_recode = case_when(
    amr$province == "Vieniane Capital" ~ "Vientiane Capital",
    amr$province == "Oudomxai" ~ "Oudomxay",
    amr$province == "Baukeo" ~ "Bokeo",
    amr$province == "Huaphan" ~ "Huaphanh",
    amr$province == "Xaiyabuly" ~ "Xayabury",
    amr$province == "Baulikhamxai" ~ "Borikhamxay",
    amr$province == "Khammuan" ~ "Khammuane",
    amr$province == "Saravan" ~ "Salavan",
    amr$province == "Champasak" ~ "Champasack",
    amr$province == "Savannakhet" ~ "Savanakhet",
    amr$province == "Xaysomboon" ~ "Xaysomboun",
    amr$province == "Xiengkhuang" ~ "Xieng Khuang",
    amr$province == "Luangnamtha" ~ "Luang Namtha",
    amr$province == "Luangprabang" ~ "Luang Prabang",
    TRUE ~ amr$province)) %>%
  select(-province) %>%
  rename(province = province_recode)
```

Recode method of collection.
```{r}
amr$spec_method [amr$spec_method  == "Haemoculture"] <- "Blood culture"
```


### Step 1

Start: The `amr` dataset contains `r nrow(amr)` rows with `r length(unique(amr$patient_id))` patients  and `r length(unique(amr$spec_id))` specimens.

Exclude the `r sum(str_detect(amr$patient_name, "LACORIS"))` rows with `patient_name` containing the character string "LACORIS".

```{r}
amr <- amr %>%
  filter(!str_detect(amr$patient_name, "LACORIS")) %>%
  select(-patient_name)
```

**The `amr` dataset contains `r nrow(amr)` rows with `r length(unique(amr$patient_id))` patients  and `r length(unique(amr$spec_id))` specimens.**

### Step 2

Start: The `amr` dataset contains `r nrow(amr)` rows with `r length(unique(amr$patient_id))` patients  and `r length(unique(amr$spec_id))` specimens.

Include only clinically significant organisms, for which the column include is set to "Yes".

```{r}
organisms <- org_include %>% 
  filter(org_include$include == "Yes") %>% 
  pull(organisms)

amr <- amr %>% 
  filter(amr$org_name %in% organisms)
```

**The `amr` dataset contains `r nrow(amr)` rows with `r length(unique(amr$patient_id))` patients  and `r length(unique(amr$spec_id))` specimens.**


### Step 3

Start: The `amr` dataset contains `r nrow(amr)` rows with `r length(unique(amr$patient_id))` patients  and `r length(unique(amr$spec_id))` specimens.

Exclude all elements for which the method is "Clotted Blood"" or "EDAT Blood".

```{r}
amr <- amr %>% 
  filter(!amr$spec_method %in% c("Clotted blood", "EDTA blood"))
```

**End: The `amr` dataset contains `r nrow(amr)` rows with `r length(unique(amr$patient_id))` patients  and `r length(unique(amr$spec_id))` specimens.**

### Step 4

Start: The `amr` dataset contains `r nrow(amr)` rows with `r length(unique(amr$patient_id))` patients  and `r length(unique(amr$spec_id))` specimens.

Exclude the rows with the following comments: "Mixed environemental gram negative organism and Staphylococcus coagulase negative", "Normal oral flora", "Normal skin flora".

```{r}
excluding_comments <- c("Mixed environemental gram negative organism and Staphylococcus coagulase negative", "Normal oral flora", "Normal skin flora")
amr <- amr %>% filter(!amr$culturecomment %in% excluding_comments, 
                      !amr$culturecomment2 %in% excluding_comments)
```

**End: The `amr` dataset contains `r nrow(amr)` rows with `r length(unique(amr$patient_id))` patients  and `r length(unique(amr$spec_id))` specimens.**

### Step 5

Start: The `amr` dataset contains `r nrow(amr)` rows with `r length(unique(amr$patient_id))` patients  and `r length(unique(amr$spec_id))` specimens.

Exclude organisms that are non signicant by following Talamlee recommendation: *"If there is an organism other than No growth in ORG_NAME and a comment of contaminant, probably contaminant or uncertain significance in culturecomment and culturecomment2, then it should NOT be considered significant."*

```{r}
exclude <- amr$org_name != "No growth" & (str_detect(amr$culturecomment, "Probable contaminant") | str_detect(amr$culturecomment, "Uncertain clinical significance") | str_detect(amr$culturecomment2, "Probable contaminant") | str_detect(amr$culturecomment2, "Uncertain clinical significance"))

amr <- amr[!exclude, ]
```

**End: The `amr` dataset contains `r nrow(amr)` rows with `r length(unique(amr$patient_id))` patients  and `r length(unique(amr$spec_id))` specimens.**

Comments in `amr` dataset `culturecomment` and `culturecomment2`:

```{r}
sort(unique(c(amr$culturecomment, amr$culturecomment2)))
```


### Step 6

Start: The `amr` dataset contains `r nrow(amr)` rows with `r length(unique(amr$patient_id))` patients  and `r length(unique(amr$spec_id))` specimens.

Exclude organisms *"Staphylococcus species - not Staph. aureus"*

```{r}
exclude <- amr$org_name == "Staphylococcus species - not Staph. aureus"

amr <- amr[!exclude, ]
```

**End: The `amr` dataset contains `r nrow(amr)` rows with `r length(unique(amr$patient_id))` patients  and `r length(unique(amr$spec_id))` specimens.**



## Validation and Quality Check

Please review the following:

```{r}
# Missing method of collection
ifelse(any(is.na(amr$spec_method)), 
       "There are rows with a missing specimen collection method",
       "Okay: all rows have a collection method"
       )

# All codes of antibiotics have a matching code in the dictionary
ifelse(is.na(setdiff(amr$antibiotic_code, antibio_codes$antibiotic_code)),
       "Okay: all codes of antibiotics have a matching code in the dictionary", 
       {print("The following antibiotic codes have no matching names:"); setdiff(amr$antibiotic_code, antibio_codes$antibiotic_code)})

# All codes of specimen types have a matching code in the dictionary
ifelse(length(setdiff(amr$spec_type, spec_codes$spec_type)) == 0,
       "Okay: all codes of specimen types have a matching code in the dictionary",
       {print("The following specimens codes have no matching names:"); setdiff(amr$spec_type, spec_codes$spec_type)})
```


## Generate Dataset for the Dashboard

Data file saved under the name: `r paste0("AMR_data_LOMWRU", "_", format(Sys.time(), "%F"), ".RData")`

```{r}
amr <- amr %>% 
  select("spec_id", "patient_id", "gender", "province", "spec_type", "spec_year",
         "spec_date", "age_years", "age_months", "age_days", "location", 
         "org_name", "esbl", "antibiotic_code", "resistance", "method", 
         "antibiotic_name", "spec_method")

data <- list(source_data = "LOMWRU",
             date_generation = Sys.Date(), 
             amr = amr)

save(list = "data", file = paste0("./data/AMR_data_LOMWRU", "_", time_start, ".RData"))
```
